\documentclass{article}
\usepackage{hyperref}
\usepackage[nohyphen]{underscore}
\usepackage[english]{babel}
\newcommand{\code}[1]{\texttt{#1}}
\begin{document}
\title{Modular Wayland compositors with wlroots}
\author{Drew DeVault et al
    \thanks{Complete list available at \url{https://github.com/swaywm/wlroots/graphs/contributors}}}
\maketitle

\begin{abstract}
wlroots is a flexible and modular library upon which Wayland compositors can be
built. It is designed to maximize several factors: code reusability,
interoperability between Wayland compositors, and flexibility for use in novel
compositor designs.
\end{abstract}

\newpage
\tableofcontents

\newpage
\section{Background}\label{background}

Wayland is a protocol used to communicate between user applications and a
\emph{compositor}. The compositor has ownership over input and output
resources-- typically displays connected over DVI, HDMI, etc, and keyboards,
mice, etc.\\

wlroots (stylized in lowercase, or as "wlr") is a subproject of the Sway
compositor. Originally, Sway was written on top of a similar library called wlc.
This library offered abstractions on top of DRM, GBM, GLES, libinput, and so
on-- allowing Sway to focus on building the software specific to the compositor.
However, with time it became clear that wlc's design is too limiting, and many
of Sway's long-term goals were difficult to achieve based on it.\\

As a result, we built wlroots. We believe it offers a substantially better
basis for development of Wayland compositors than any competing approach.
We offer a powerful library that removes most of the heavy lifting from building
a Wayland compositor while providing compositor authors as much flexibility in
their design as they might gain by doing it all themselves-- I've jokingly
referred to wlroots as
"about 30,000 lines of code you were going to write anyway." We hope that
compositors based on wlroots will be written faster, work better, and will
be better citizens of the broader Wayland ecosystem.

\subsection{Alternatives}\label{alternatives}

Several alternative approaches to building compositors exist.\\

\textbf{libweston} is the most notable alternative. It is a subproject of the
Weston compositor, the reference Wayland compositor. However, it is generally
poorly suited to novel compositors that don't use the traditional desktop
paradigm. It is also less modular and difficult to break into several pieces.
Additionally, as a consequence of being refactored out of an existing
compositor, it inherits many design decisions of Weston itself, which may not be
amenable to other compositors.

\textbf{swc} and \textbf{wlc} are similar. They both strike a reasonable balance
between size and features, but both make decisions for the compositor that lock
them into particular designs and make novel compositors more difficult to
write. The advantage of their approach is that it is easy to use them to build a
simple working compositor - in a matter of minutes, not days or weeks.

Another approach is to simply write your compositor entirely in-house. This is
the approach favored by \textbf{mutter} and \textbf{kwin} of the GNOME and KDE
projects respectively. The advantage of this approach is the flexibility to
design your compositor in any way you please. However, lots of work is
duplicated between projects with this approach, which can (and does) lead to
compositors which are subtly incompatible with each other.

The approach taken by \textbf{wlroots} offers reusability and consistency
between compositors based on it, and offloads lots of work from the compositor
implementation. This is accomplished without making design decisions on behalf
of the compositor implementation: we believe virtually any novel kind of
compositor can be built on top of wlroots.

The main disadvantage of our approach is that batteries are not included. More
work is required to get a working compositor compared to libweston or wlc.
However, you will not hit the limitations of these libraries, and will more
quickly arrive at a diverse featureset than with the kwin or mutter approaches.

\section{High-level Design}\label{highlevel}

Let's summarize the important high-level components of wlroots.

\subsection{Backends}

The wlroots backends are similar to backends in other compositors, but are more
easily used standalone or in composite. A compositor based on wlroots can use
zero or more backends at a time, dynamically adding and removing different
backends (and different types of backends) at runtime. Each backend is
responsible for providing zero or more input and output devices, as well as an
EGL context (and support for maneuvering between EGL contexts is provided). This
design enables us to support use cases like multiple GPUs, scripting inputs on
your desktop by adding a headless backend, temporarily adding an RDP backend
with mirrored outputs to your session, and so on.

We currently offer DRM, libinput, Wayland, X11, and headless backends. We have
also written a virtual "multi" backend which acts as a container for multiple
backends, with which you may add or remove backends at runtime. Use of this
multi-backend is necessary in most non-trivial compositors, for example, to use
DRM and libinput at the same time. Also provided is an interface with which
compositor authors can create their own backend implementations for more novel
systems, and we plan to write an RDP and fbdev backends in the future.

\subsection{Shells}

Several implementations of Wayland shell state machines are included in wlroots.
Their purpose is to provide largely non-opinionated implementations of the
various shells available on a Wayland desktop, and to delegate to the compositor
for when opinions are necessary. We do not provide a grand unified shell
abstraction. Rather, we expose the subtleties of each shell to the compositor
author so they may have finer control over how these subtleties are addressed
given the constraints of their particular compositor design.

Provided are implementations of wl_shell, xdg_shell, and Xwayland. The latter is
quite thorough, on the assumption that the last thing the author of a Wayland
compositor wants to do is write an X11 window manager too. However, it is as
flexible as any other shell, and you are able to reach in and tweak it to suit
your compositor's needs.

\subsection{Wayland Protocols}

In addition to shell protocols, implementations of various other protocols are
provided. This includes core protocols like wl_data_device_manager,
wl_compositor, wl_surface, and so on, but also includes some third-party
protocols such as Orbital's screenshooter protocol and GTK's primary selection
protocol. When appropriate, integrations with wlroots implementations of other
protocols may be necessary, such as interactions between wlr_seat and
wlr_data_device_manager, or syncronizing the GTK's Wayland primary selection
with Xwayland.

Each of these protocols, like the shells and everything else, is opt-in. The
compositor author chooses if our implementations suit their needs, and if not,
their own implementation is easily used instead and interops correctly with the
rest of wlroots. For some important protocols (such as wlr_seat), disentangling
it from the rest of wlroots can be more difficult-- but in these cases we've
made our implementation extremely flexible to make it less likely that it will
not suit your needs.

\subsection{Convenience Utilities}

We also provide optional implementations of functionality common to most Wayland
compositors, which can be integrated with the rest of wlroots to any degree the
compositor author chooses. This includes modules which help you manage the
layout of outputs in physical space (and arranging windows on a virtual
desktop), help you map input from pointers, touch, etc onto cursors shown
on-screen, or to read cursor bitmaps from Xcursor themes for use when your
compositor overrides client cursors.

These interfaces are also quite modular and composable, and will generally work
in the absence of other wlroots features, or offer more functionality in their
presence. The wlr_cursor module, for example, will allow you to limit cursor
movement to within the bounds of a wlr_output_layout, should you choose to
utilize that as well. We also provide a renderer module, which utilizes gles2 to
provides simple rendering support which should be suitable for any compositor
that does not require complex shaders or projections.

\subsection{Rootston}

Bringing all of this together is our reference compositor, Rootston. As Weston
is to libweston, Rootston is to wlroots. However, Rootston is more conservative -
it is a very simple stacking window manager that has only a few convenience
features on top of demonstrating the capabilities of wlroots. This makes it
quite unsuitable as a practical end-user compositor. However, in its role of
demonstrating the capabilities of wlroots, it does a fine job, and any
compositor author hoping to use wlroots will likely find themselves reading, and
perhaps copying, large swaths of Rootston's code.

Like the wlroots library itself, Rootston is MIT licensed and users are
encouraged to copy and adapt any code they find useful for their own
compositors. Many of the batteries that are "not included" with wlroots can be
found in Rootston.

\section{Promoting an Interoperable Ecosystem}

We believe that it is tremendously important for more Wayland compositors to be
extensible by their users and for these tools to work across the entire Wayland
desktop ecosystem. No one will switch from X11 to Wayland for security alone--
we must make compositors that offer compelling features and do not ask the user
to sacrifice the workflows they've grown used to, insecure as they may have been
on X11.

We are committed to implementing many Wayland protocol extensions to support
these features. Where extensions already exist, we are implementing them: we
provide implementations of KDE's server decoration protocol, GTK's primary
selection protocol, and Orbital's screenshot protocol; and in the future we
intend to implement more. Where protocols do not exist, we are creating new
ones.

Consider an on-screen keyboard-- getting it right is complicated. There are
many keyboard layouts to consider, and localization may be difficult. Support
for typing in pinyin (Chinese) or romaji (Japanese) will be difficult to do
correctly. Useful features like providing prompts for autocomplete will also
take significant developer investment. We could do this many times for many
Wayland compositors, or we could push for a single implementation that is
compatible with all of our software.

The state of interopability in the Wayland ecosystem is something the wlroots
team is deeply concerned about, and we worry that many compositors are not
taking this problem seriously. This is why we are implementing the protocols of
other desktops and toolkits, and building our own when these prove insufficient.
The first of our new protocols is surface layers. We believe this protocol can
support numerous use-cases in a single simple protocol:

\begin{itemize}
    \item desktop shells
    \item lock screens
    \item on screen keyboards
    \item notification daemons
    \item custom program launchers
    \item magnifying glass tools
\end{itemize}

The details of the protocol or its implementation are out of scope for this
document, and in any case are changing as we develop it. Here it serves as a
call to action for other compositors to participate in its development and
implementation, and an example of the sorts of functionality we're committed to
providing on Wayland.

\subsection{Security Considerations}

A major concern in implementing many of these features is doing them in a
secure way. Making these interfaces available to third-party software securely
is the subject of ongoing research and discussion, but we do not believe that
the issue is too poorly understood to proceed with implementing these features.
In wlroots, we have chosen to leave securing these protocols as a matter for the
compositor to support via wl_display_set_global_filter.

Let's summarize ongoing efforts by compositors in this area. The Sway compositor
has, as of writing, invested the most effort in making privledged protocols
available to third parties. This is not without consequences-- Sway's current
implementation has known flaws. We are comfortable being more liberal in our
exploration in this area, hoping that we can become more secure over time. Our
justification for this approach is that, at worst, we begin with a system no
more secure than X11, but over time we will be able to secure it more as we
experiment with different approaches.

Sway's approach today is configuration files, owned by root, which contain a
list of paths to executables and the permissions granted to each. To enforce
these, we look them up via /proc/[pid]/exe. This has a number of drawbacks we
are working to address. First, this is not very effective for interpreted
programs, e.g. /usr/bin/python3. Additionally, it may be possible on some
platforms and in some conditions to manipulate the value of /proc/[pid]/exe.
Utilizing procfs is also not portable.

In the future we're exploring the possibility of only permitting secure access
for clients which were forked from Sway. This approach should solve most of the
outstanding problems, but other deficiencies remain. Granting permissions to
previously un-approved clients at runtime (with user content) requires
additional work (something the Orbital compositor has explored). Offering
similar permissions to children of those processes when appropriate is also a
non-trivial problem.

Collaboration with the broader Wayland community in the matter of security is
highly sought after. We ask you to participate in our discussions on the
subject; contact information is available at the end of this paper.

\section{Project Status}

Integrations with wlroots are well underway in the Wayland community.\\

\textbf{Sway} is a Wayland compositor compatible with the i3 window manager and
is the motivating force behind wlroots, and our port from wlc to wlroots is
going well. We've found it most effective to set aside our current code and
start from scratch, importing large parts of the old code as it becomes
necessary. Porting is no simple process, but it seems that it will be well worth
it. Dozens of features that our users have been asking for have been blocked
behind this migration for a long time, and many are already supported in our
port\footnote{Enumerated here:
\url{https://github.com/swaywm/sway/issues/1524}}.\\

\textbf{Way-Cooler} is also working on porting their compositor from wlc to
wlroots. In order to do so, they are developing \textbf{wlroots-rs}, a Rust
wrapper over the wlroots API. They intend to utilize this to implement a large
subset of the Awesome window manager's Lua API, providing a compatible
alternative to AwesomeWM in a similar fashion as Sway does for i3. Their first
order of business is providing an equivalent of Rootston written in Rust.\\

\textbf{waymonad} is a new compositor based on wlroots-- not a port of another
compositor. Like Sway fills i3's niche and Way-Cooler fills AwesomeWM's niche,
waymonad intends to fill Xmonad's niche (though it seems that perfect
compatibility is not presently a concern). To this end \textbf{hsroots} is
underway, Haskell wrappings for the wlroots API.\\

We've heard murmurings of a compositor written in Go as well. We are excited
about and prepared to support any other compositors that want to port their
plumbing to wlroots, as well as any new compositors that would like to use
wlroots from the outset. As the Sway port and other projects make progress, we
continue to prove the effecacy of the wlroots approach and stabilize its API,
and we are pushing towards 1.0 soon. Please reach out if you have feedback or
wish to participate.\\

\url{https://github.com/swaywm/wlroots}\\

\url{irc://irc.freenode.net/#sway-devel}

\end{document}
